---
title: "Assignment 5B"
author: "Khandker Qaiduzzaman"
format: html
editor: visual
---

# Objective

Calculate each chess player’s expected score using the ELO rating system based on the provided dataset, compare it with their actual score, and identify the top five players who over-performed and under-performed relative to expectations.

# Approach

### Dataset description

For this assignment, I will use the output dataset generated from Project 1, which is a preprocessed CSV file containing structured chess tournament results. The dataset can be accessed here: <https://raw.githubusercontent.com/NafeesKhandker/ELO-Calculations/refs/heads/main/Chess_result_data.csv>

This dataset is already cleaned and organized, making it suitable for direct analysis. The dataset contains the following columns:

-   **pair_num**: Unique pairing number assigned to each player

-   **Player_name**: Name of the player

-   **State**: Player's stat

-   **Total_points**: Total points scored by the player in the tournament

-   **Rating**: Player’s pre-tournament rating

-   **Avg_opponent_rating**: Average rating of all opponents faced by the player

This structured dataset simplifies the calculation of expected scores and performance differences using the ELO rating system.

### Anticipated Challenges

The main anticipated challenge for this project is implementing the ELO expected score calculation correctly and efficiently. Since each player competes against multiple opponents, the expected score must be calculated individually for each match and then aggregated. Another challenge is ensuring accurate numerical computation, especially when dealing with floating-point precision during exponential calculations.

### Mock Output

Pair_num, Player_name, State, Actual_score, Expected_score, Difference

1, GARY HUA, ON, 6.0, 5.2, +0.8

2, DAKSHESH DARURI, MI, 6.0, 4.3, +1.7

3, ADITYA BAJAJ, MI, 6.0, 1.8, +4.2

4, PATRICK H SCHILLING, MI, 5.5, 4.9, +0.6

5, HANSHI ZUO, MI, 5.5, 5.0, +0.5

### Implementation of ELO calculation

```{r}
library(tidyverse)
library(gt)

url <- "https://raw.githubusercontent.com/NafeesKhandker/ELO-Calculations/refs/heads/main/Chess_result_data.csv"

df <- read_csv(file = url, show_col_types = FALSE, progress = FALSE)

df |> gt()
```

In this step, we are calculating the expected score per game for a player. The expected score for Player A against Player B (average opponents rating in our case) is given by:

$$
E_A = \frac{1}{1 + 10^{\frac{R_B - R_A}{400}}}
$$

Where:

-   $E_A$ = Expected score of Player A\
-   $R_A$ = Rating of Player A\
-   $R_B$ = Rating of Player B

The second Elo formula updates a player's rating after each game based on the actual and expected scores:

$$
R_{\text{new}} = R_{\text{old}} + K \times (S - E)
$$

Where:

-   $R_old$ = current rating of the player\
-   $R_new$ = updated rating after the game\
-   $K$ = K-factor (adjustment coefficient, e.g., 20, 32)\
-   $S$ = actual score of the game (1 = win, 0.5 = draw, 0 = loss)\
-   $E$ = expected score, calculated as:\

```{r}
library(tidyverse)

# Set K-factor (adjust if needed)
K <- 32

df_final <- df %>%
  mutate(
    # Ensure numeric
    rating_pre = as.numeric(rating_pre),
    opponent_rating = as.numeric(opponent_rating),

    # 1. Expected Score (Elo Formula)
    expected_score = 1 / (1 + 10 ^ ((opponent_rating - rating_pre) / 400)),

    # 2. Actual Score
    actual_score = case_when(
      result_type == "W" ~ 1,
      result_type == "D" ~ 0.5,
      result_type == "L" ~ 0
    ),

    # 3. Rating Change (Second Elo Formula)
    rating_change = K * (actual_score - expected_score),

    # 4. Updated Rating per game
    updated_rating = rating_pre + rating_change
  )

df_final |> gt()
```

In the following code, I aggregated the data to show total expected and total actual score in adjacent columns. Also, I am aggregating the average expected and average actual scores. Finally, the performance difference is calculated which is the target column of this assignment.

```{r}
# -----------------------------------------
# Aggregate per player
# -----------------------------------------

df_summary <- df_final %>%
  group_by(pair_num, player_name, state, rating_pre) %>%
  summarise(
    games = n(),

    # Expected vs Actual
    total_expected = sum(expected_score, na.rm = TRUE),
    total_actual = sum(actual_score, na.rm = TRUE),

    avg_expected = mean(expected_score, na.rm = TRUE),
    avg_actual = mean(actual_score, na.rm = TRUE),

    # Performance Difference
    performance_diff = total_actual - total_expected,

    # Rating Changes
    total_rating_change = sum(rating_change, na.rm = TRUE),
    final_rating = first(rating_pre) + total_rating_change,

    .groups = "drop"
  ) %>%
  arrange(pair_num)

# View results
df_summary |> gt()
```

Top 5 overperformers relative to their expected score is shown below:

```{r}
# Top 5 Overperformers
top_overperformers <- df_summary %>%
  arrange(desc(performance_diff)) %>%
  slice_head(n = 5) %>%
  select(player_name, state, total_actual, total_expected, performance_diff, final_rating)

# Print results
top_overperformers |> gt()
```

Top 5 Underperformers relative to their expected score is shown below:

```{r}
# Top 5 Underperformers
top_underperformers <- df_summary %>%
  arrange(performance_diff) %>%
  slice_head(n = 5) %>%
  select(player_name, state, total_actual, total_expected, performance_diff, final_rating)

# Print results
top_underperformers |> gt()
```

### Conclusion

In this analysis, we calculated each player’s expected scores using the Elo rating system and compared them to their actual performance. The results highlighted which players overperformed or underperformed relative to expectations, providing insight into performance consistency. By quantifying performance differences, this approach allows for a fair evaluation of players’ relative strengths and weaknesses. Overall, the Elo-based analysis demonstrates an effective method to predict and assess player outcomes in competitive settings.

### References

1.  Solon, N. (2022, December 12). How Elo ratings actually work. Chess.com. https://www.chess.com/blog/NateSolon/how-elo-ratings-actually-work
2.  Grime, J. (2019, February 15). *The Elo rating system for chess and beyond* \[Video\]. YouTube. [https://www.youtube.com/watch?v=AsYfbmp0To0](https://www.youtube.com/watch?v=AsYfbmp0To0&utm_source=chatgpt.com)
3.  OpenAI. (2026, March 1). *ChatGPT conversation on ELO ratings, R analysis, and data visualization* \[AI-generated transcript\]. OpenAI. <https://chat.openai.com/>
